
# Aggregating

In each of the examples, the expression aggregates over all the rows, and returns a single row.  

That's what makes these expressions **aggregate expressions**. They can combine values from multiple rows, aggregating them together. These are different from the arithmetic operations like `+`, `-`, `*` and `/`, or functions like `ROUND`, `STRSUB` , which operate independently on the values in each row, and return a column whose length equals that of the whole table. Those are called **non-aggregate** expressions. 

One need to be careful about mixing aggregate and non-aggregate operations. You can use aggregate and non-aggregate operations together in an expression as in the following examples.    

```{sql connection=con}
SELECT ROUND(AVG(salary), 1) AS avg_salary 
FROM employees 
```

```{sql connection=con}
SELECT SUM(salary * 0.5) AS avg_half_salary 
FROM employees 
```

Unfortunately, there are also invalid mix of aggregation expression and non-aggreagate expressions. Suppose we want to know the difference between one's salary and the highest salary in the table

```{sql connection=con, eval = FALSE}
-- not run
SELECT salary - max(salary) 
FROM employees
```

This query will throw an error message like **"employees.salary must appear in the GROUP BY clause or be used in an aggregate function"**. For R users, this may be a bit confusing, since similar dplyr expressions work just fine:  

```{r}
library(dplyr)
employees <- readr::read_csv("data/employees.csv")

employees %>%
  mutate(difference = salary - max(salary))
```

This is because the R language recycles the one element scalar `mean(salary)` to match its length to `salary`. SQL, on the other hand, does not know how to subtract one row from multiple rows. 

For now, the workaround is to use **subqueries**. Later we will see another solution using window functions in Chapter \@ref(window-functions).


```{sql connection=con}
SELECT salary - 
  (SELECT MAX(salary) FROM employees)
  AS difference
FROM employees
```


Also, you cannot select aggregating expressions and regular columns in parallel. For example, the following query has two items in the `SELECT` list. 

```{sql connection=con, eval = FALSE}
-- not run
SELECT first_name, max(salary) FROM employees
```

The first is just the column reference first name, that evaluates as a scalar expression. It returns a value for each row in the `employees` table. The second is the aggregate expression, max salary. That returns one row that aggregates all the salary values in the `employees` table.




### rowwise aggregating

There are functions in SQL that are designed to perform computations per row, instead of across rows. For them, we need not to worry about the unmatched length problem. 

For example, there are non-aggregate `GREATEST()` for `MAX()`, and `LEAST` for `MIN()`

```{sql connection=con}
-- what is the dominant and least-used color for each crayon? 
SELECT 
  red, green, blue, GREATEST(red, green, blue), LEAST(red, green, blue)
FROM crayons
```


This is much like the rowwise mechanism introduced since dplyr 1.0 

```{r}
crayon <- readr::read_csv("data/crayons.csv")
crayon %>%
  rowwise() %>% 
  mutate(dominant = max(c(red, green, blue)),
         `least used` = min(c(red, green, blue)))
```



## The `GROUP BY` clause. 

No surprise, the `GROUP BY` clasue works hand in hand with aggregate expressions, as `group_by()` does with `summarize()` in dplyr.  

```{sql connection=con}
SELECT min_age, COUNT(*)
FROM games
WHERE list_price > 10
GROUP BY min_age;
```

Grouping expressions using column aliases, availabe in PostgreSQL, Impala, Mysql. 

```{sql connection=con}
SELECT list_price < 10 AS low_price, count(*) 
FROM games
GROUP BY low_price
```

```{sql connection=con}
SELECT list_price < 10, COUNT(*) FROM games GROUP BY list_price < 10;
```

```{sql connection=con}
SELECT (list_price) > 20 AS over_20, max_players, COUNT(*)
FROM games
GROUP BY over_20, max_players;
```
```{sql connection=con}
SELECT * FROM games
```



One of the major limitations of the `GROUP BY` clause is that it left you few choices in the `SELECT` list: **either an aggregate expression or columns that appear in `GROUP BY`**  


## Common aggregating functions



## NULL values in aggregation  

The `COUNT` function was designed to be consistent with these other aggregate functions
except in the case where you use `COUNT(*)`. So the general rule is that aggregate functions ignore `NULL` values, and the one exception to that rule is when you use `COUNT(*)`.   


With some SQL engines, you can specify more than one column reference or expression after the `DISTINCT` keyword in the `COUNT` function. This returns the number of unique combinations of the specified columns that exist in the data. This works in Hive, Impala, and MySQL, but **not** in PostgreSQL. 


```{sql connection=con, eval = FALSE}
-- this won't work in PostgreSQL
SELECT COUNT(DISTINCT red, green, blue) FROM crayons
```

Also, with some SQL engines, you can use more than one `COUNT(DISTINCT)` in a `SELECT` list,
like in this example which uses the crayons table.

```{sql connection=con}
-- PostgreSQL allows for multiple COUNT(DISTINCT) in one select list
SELECT 
  pack,
  COUNT(DISTINCT red) AS red_count,
  COUNT(DISTINCT green) AS green_count,
  COUNT(DISTINCT blue) AS blue_count
FROM crayons
GROUP BY pack
```

Here, the COUNT function is used three separate times in the `SELECT` list and the `DISTINCT` keyword is included in all three. So the result set has three columns giving the number of unique values in the red column, the number of unique values in the green column, and the number of unique values in the blue column. But with some other SQL engines including Impala, you are limited to only one` COUNT(DISTINCT)` per `SELECT` list. 




## Using `GROUP BY` in non-aggregating scenarios 


We may question the data quality of `us_counties_2010` cencus: specifically, can there be repetition of observations on counties on the state level? We may want to use `DISTINCT(geo_name)` and `GROUP BY state_us_abbreviation` to eliminate those repetitions:  

```{sql connection=con, eval = FALSE}
-- not run
SELECT DISTINCT(geo_name) 
FROM us_counties_2010
GROUP BY state_us_abbreviation
```

This works in dplyr 

```{r}
# dplyr
us_counties_2010 <- readr::read_csv("data/us_counties_2010.csv")
us_counties_2010 %>% 
  group_by(state_us_abbreviation) %>% 
  distinct(geo_name)
```






