
# Aggregating

In each of the examples, the expression aggregates over all the rows, and returns a single row.  

That's what makes these expressions **aggregate expressions**. They can combine values from multiple rows, aggregating them together. These are different from the arithmetic operations like `+`, `-`, `*` and `/`, or functions like `ROUND`, `STRSUB` , which operate independently on the values in each row, and return a column whose length equals that of the whole table. Those are called **non-aggregate** expressions. 

One need to be careful about mixing aggregate and non-aggregate operations. You can use aggregate and non-aggregate operations together in an expression as in the following examples.    

```{sql connection=con}
SELECT ROUND(AVG(salary), 1) AS avg_salary 
FROM employees 
```

```{sql connection=con}
SELECT SUM(salary * 0.5) AS avg_half_salary 
FROM employees 
```

Unfortunately, there are also invalid mix of aggregation expression and non-aggreagate expressions. Suppose we want to know the difference between one's salary and the highest salary in the table

```{sql connection=con, eval = FALSE}
-- not run
SELECT salary - max(salary) 
FROM employees
```

This query will throw an error message like **"employees.salary must appear in the GROUP BY clause or be used in an aggregate function"**. For R users, this may be a bit confusing, since similar dplyr expressions work just fine:  

```{r}
library(dplyr)
employees <- readr::read_csv("data/employees.csv")

employees %>%
  mutate(difference = salary - max(salary))
```

This is because the R language recycles the one element scalar `mean(salary)` to match its length to `salary`. SQL, on the other hand, does not know how to subtract one row from multiple rows. 

For now, the workaround is to use **subqueries**. Later we will see another solution using window functions in Chapter \@ref(window-functions).


```{sql connection=con}
SELECT salary - 
  (SELECT MAX(salary) FROM employees)
  AS difference
FROM employees
```


Also, you cannot select aggregating expressions and regular columns in parallel. For example, the following query has two items in the `SELECT` list. 

```{sql connection=con, eval = FALSE}
-- not run
SELECT first_name, max(salary) 
FROM employees
```

The first is just the column reference first name, that evaluates as a scalar expression. It returns a value for each row in the `employees` table. The second is the aggregate expression, max salary. That returns one row that aggregates all the salary values in the `employees` table.


## Aggregate Functions


There common aggregate functions for summary statistics like `MIN()`, `MAX()`, `AVG`, `SUM`.  


Additionally, there are two aggregate functions that do not perform direct computation based on values of columns.

`DISTINCT` can be used as keyword and functions alike. 

`COUNT`  

- `COUNT(*)`: Returns total number of records 

- `COUNT(column_name)`: Return number of **Non Null** values over the column salary. i.e 5.

- `COUNT(DISTINCT Salary)`:  Return number of **distinct Non Null** values over the column salary .i.e 4


`CASE WHEN` is often used in calculating proportions and percentages, the standard syntax is 



## Rowwise Aggregation

There are functions in SQL that are designed to perform computations per row, instead of across rows. For them, we need not to worry about the unmatched length problem. 

For example, there are non-aggregate `GREATEST()` for `MAX()`, and `LEAST` for `MIN()`

```{sql connection=con}
-- what is the dominant and least-used color for each crayon? 
SELECT 
  red, green, blue, 
  GREATEST(red, green, blue), 
  LEAST(red, green, blue)
FROM crayons
```


This is much like the rowwise mechanism introduced since dplyr 1.0 

```{r}
crayon <- readr::read_csv("data/crayons.csv")
crayon %>%
  rowwise() %>% 
  mutate(dominant = max(c(red, green, blue)),
         `least used` = min(c(red, green, blue)))
```



## The `GROUP BY` Clause

No surprise, the `GROUP BY` clasue works hand in hand with aggregate expressions, as `group_by()` does with `summarize()` in dplyr.  

```{sql connection=con}
SELECT min_age, COUNT(*)
FROM games
WHERE list_price > 10
GROUP BY min_age;
```

Grouping expressions using column aliases, available in PostgreSQL, Impala, Mysql. 

```{sql connection=con}
SELECT list_price < 10 AS low_price, count(*) 
FROM games
GROUP BY low_price
```

```{sql connection=con}
SELECT list_price < 10, COUNT(*) FROM games GROUP BY list_price < 10;
```

```{sql connection=con}
SELECT (list_price) > 20 AS over_20, max_players, COUNT(*)
FROM games
GROUP BY over_20, max_players;
```




One of the major limitations of the `GROUP BY` clause is that it left you few choices in the `SELECT` list: **either an aggregate expression or columns that appear in `GROUP BY`**  


## Common Aggregating Functions



## `NULL` in Aggregation  

The `COUNT` function was designed to be consistent with these other aggregate functions
except in the case where you use `COUNT(*)`. So the general rule is that aggregate functions ignore `NULL` values, and the one exception to that rule is when you use `COUNT(*)`.   


With some SQL engines, you can specify more than one column reference or expression after the `DISTINCT` keyword in the `COUNT` function. This returns the number of unique combinations of the specified columns that exist in the data. This works in Hive, Impala, and MySQL, but **not** in PostgreSQL. 


```{sql connection=con, eval = FALSE}
-- this won't work in PostgreSQL
SELECT COUNT(DISTINCT red, green, blue) F
ROM crayons
```

Also, with some SQL engines, you can use more than one `COUNT(DISTINCT)` in a `SELECT` list, like in this example which uses the crayons table.

```{sql connection=con}
-- PostgreSQL allows for multiple COUNT(DISTINCT) in one select list
SELECT 
  pack,
  COUNT(DISTINCT red) AS red_count,
  COUNT(DISTINCT green) AS green_count,
  COUNT(DISTINCT blue) AS blue_count
FROM crayons
GROUP BY pack
```

Here, the `COUNT` function is used three separate times in the `SELECT` list and the `DISTINCT` keyword is included in all three. So the result set has three columns giving the number of unique values in the red column, the number of unique values in the green column, and the number of unique values in the blue column. But with some other SQL engines including Impala, you are limited to only one` COUNT(DISTINCT)` per `SELECT` list. 




## Find distinct combinations 


Suppose we want to find distinct combinations of state and county in the census (which should not be repeated). `GROUP BY ` can be used for this sort of task. 

```{sql connection=con}
SELECT state_us_abbreviation, geo_name
FROM us_counties_2010
GROUP BY state_us_abbreviation, geo_name
```





Unlike in dplyr, `distinct` cannot be used in conjunction with `group by ` in SQL 

```{r}
# this works in 
us_counties_2010 <- readr::read_csv("data/us_counties_2010.csv")
us_counties_2010 %>% 
  group_by(state_us_abbreviation) %>% 
  distinct(geo_name)
```

```{sql connection=con, eval = FALSE}
-- sql does not have a "grouped" distinct
-- not run
SELECT DISTINCT(geo_name) 
FROM us_counties_2010
GROUP BY state_us_abbreviation
```



A workaround is using the `array` data type (Section \@ref(array))

```{sql connection=con}
SELECT  state_us_abbreviation, ARRAY_AGG(DISTINCT geo_name)
FROM us_counties_2010
GROUP BY state_us_abbreviation
LIMIT 1 
```

## Percentages 

A common used of `CASE` is to compute percentage or proportion. And the trick is to let `CASE` do the group by job manually. Suppose we want to know each school's share in the total salary.   

```{sql connection=con}
SELECT 
  sum(CASE WHEN school = 'F.D. Roosevelt HS' THEN salary ELSE 0 END
) / (sum(salary) * 1.0) as porp_fd,
  sum(CASE WHEN school = 'Myers Middle School' THEN salary ELSE 0 END
) / (sum(salary) * 1.0) as prop_myers
FROM teachers
```

The command can be easily adapted to, say, compute relative frequency. The following SQL command calculates the relative frequency of each school's teacher

```{sql connection=con}
SELECT 
  sum(CASE WHEN school = 'F.D. Roosevelt HS' THEN 1 ELSE 0 END
) / (count(*) * 1.0) as freq_fd,
  sum(CASE WHEN school = 'Myers Middle School' THEN 1 ELSE 0 END
) / (count(*) * 1.0) as freq_myers
FROM teachers
```

Yet this requires us to know all levels of school, use repeated syntax, and results in a wide data from that is often not suitable for data analysis. So this computing strategy is only useful when care about a single level or a small number of levels. 

An alternative is using aggregated subqueries (Section \@ref(filtering-by-aggregation))

```{sql connection=con}
SELECT school, 
      sum(salary) / (max(total_salary) * 1.0) as salary_prop,
      count(*) / (max(total_count) * 1.0) as rel_freq
FROM teachers, 
    (SELECT sum(salary) as total_salary, count(*) as total_count
    FROM teachers) as t
GROUP BY school
``` 
The trick is to first create the aggregation needed (in this case, total salary and counts) with aggregating subqueries, and rely on `GROUP BY` to do the calculation. Since total salary and count are scalar columns, take their max is simply to satisfy the requirements of `GROUP BY`. 







